\LARGE \bf Portfolio Optimization Using Preference Flow Based on Statistical Arbitrage

### 2017-05-17 tex/article.tex
diplić

\usepackage{lmodern}

@@

  The task of portfolio optimization is to try to enhance various criteria, which most of the time include maximization of expected return and minimization of deviation\dots

@@

  In this paper, we present a new method that uses predictions obtained by statistical arbitrage method, with confidence as a proxy for describing the preference relations between pairs of assets.
  Next, a graph is formed based on those relations, so that in-depth analysis of assets' interaction with each other might be performed.
  Finally, assets are put into order by preference, and picked into the portfolio.
  The idea of this method is to create a generalization of statistical arbitrage methods that will be more robust and perform better when working on a larger number of assets.

@@

  Let $\Omega$ be any set of entities.
  Preference relation $\succ_{\Omega \times \Omega}$ is a \textit{strict weak ordering} that describes the way humans prefer some entity over another.
  This relation is specific in that it is ($\forall x, y, z \in \Omega$):
  \begin{itemize}
    \item \textit{irreflexive}: every entity $x$ is not preferrable over itself, %($\neg \left( x \succ x \right)$),
    \item \textit{asymmetrical}: if $x$ is preferrable over some $y$, then $y$ is not preferrable over $x$, %($x \succ y \Rightarrow \neg \left( y \succ x \right)$),
    \item \textit{transitive}: if $x$ is preferrable over $y$, and $y$ is preferrable over $z$, then $x$ is also preferrable over $z$, %($x \succ y \wedge y \succ z \Rightarrow x \succ z$),
    \item \textit{transitive in incomparability} (noting that $x$ and $y$ may be \textit{incomparable}, i.e. neither $x$ is preferrable over $y$, nor $y$ is preferrable over $x$): if $x$ is incomparable with $y$, and $y$ is incomparable with $z$, then $x$ is also incomparable with $z$.
  \end{itemize}

@@

  The measure of this preference flow is determined by a statistical arbitrage method, and it corresponds to the magnitude of a pair of assets prices ratio going out of what is considered statistically confident range.
  This is illustrated on Fig. \ref{fig:devmag}.

  The main idea of preference flow is that preference of a particular node depends on amount of preference flowing in and out of it.
  This allows for some kind of generalized statistical arbitrage over multiple assets at a time.

@@

    \item Afterwards, we can calculate the reconstruction $\matr{f}^*$ of preference  flow by simply plugging back $\matr{\phi^*}$ into (\ref{eq:flowformula}):
    \begin{equation}
    \matr{f^*} = \matr{B} \matr{\phi^*}.
    \end{equation}
    The reconstructed preference flow $\matr{f^*}$ compared to the original preference flow $\matr{f}$ may even contain some new and/or lose some old edges.
    In addition, $\matr{B}$, $\matr{\phi^*}$, and $\matr{f^*}$ now describe a consistent graph $\graph{G}^*$.
    This allows now to define a consistency measure $\kappa$ as follows:
    \begin{equation}
    \label{eq:consistency}
    \kappa = \frac{\left \lVert \matr{f^*} \right \rVert}{\left \lVert \matr{f} \right \rVert}.
    \end{equation}
    Equation (\ref{eq:consistency}) represents the cosine of the angle between $\matr{f}$ and $\matr{f^*}$ in the column space of matrix $\matr{B}$.
    Consistency measure $\kappa$ tells us how consistent graph $\graph{G}$ was, compared to the $\graph{G}^*$.
    $\kappa$ ranges from 0 to 1, with 0 being full inconsistency (virtually unreachable), and 1 meaning full consistency.

@@

  \subsection{Creating the graph of preference flow}

  Using the obtained $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, and $d_{i,j}^{(t)}$ it is now possible to create a graph of preference flow among assets for each time step $t$.
  Considering one time step $t$, we find all such pairs of assets $(i,j)$ for which holds:

  \begin{equation}
    \label{eq:thresh}
    \left| c_{i,j}^{(t)} - m_{i,j}^{(t)} \right| > \alpha \cdot d_{i,j}^{(t)},
  \end{equation}
  i.e. current log price difference is at least $\alpha$ deviations distant from mean value of the past time window.
  Parameter $\alpha$ determines how many pairs of assets should constitute the graph at current time step.

  Afterwards, for each observed pair $(i,j)$ that exceeds the threshold we add into graph vertices $i$ and $j$, with a weighed edge of weight $w_{i,j}^{(t)}$ going from $i$ to $j$. Weight $w_{i,j}^{(t)}$ is obtained as:

  \begin{equation}
    \label{eq:weight}
    w_{i,j}^{(t)} = \left. \left(c_{i,j}^{(t)} - m_{i,j}^{(t)}\right) \middle/ d_{i,j}^{(t)} \right..
  \end{equation}

@@

  \section*{APPENDIX}

  \subsection{Rolling mean and variance algorithm}
  \label{sub:rolling}

  \verb|TODO|

>>>>

%\usepackage{lmodern}

@@

%  The task of portfolio optimization is to try to enhance various criteria, which most of the time include maximization of expected return and minimization of deviation\dots

@@

  In this paper, we propose a new method that uses predictions obtained by statistical arbitrage method, with confidence as a proxy for describing the preference relations between pairs of assets.
  Next, a graph is formed based on those relations, so that in-depth analysis of assets' interaction with each other might be performed.
  Finally, assets are put into order by preference, and picked into the portfolio.
  The idea of this method is to create a generalization of statistical arbitrage methods that will be more robust and perform better when working on a larger number of assets.

@@

  Let $\Omega$ be any set of entities.
  Preference relation $\succ$ defined over $\Omega \times \Omega$ is a strict weak ordering that describes the way humans prefer some entity over another.
  This relation is specific in that it is ($\forall x, y, z \in \Omega$):
  \begin{itemize}
    \item \textit{irreflexive}: every entity $x$ is not preferrable over itself, %($\neg \left( x \succ x \right)$),
    \item \textit{asymmetrical}: if $x$ is preferrable over some $y$, then $y$ is not preferrable over $x$, %($x \succ y \Rightarrow \neg \left( y \succ x \right)$),
    \item \textit{transitive}: if $x$ is preferrable over $y$, and $y$ is preferrable over $z$, then $x$ is also preferrable over $z$, %($x \succ y \wedge y \succ z \Rightarrow x \succ z$),
    \item \textit{transitive in incomparability} (noting that $x$ and $y$ may be \textit{incomparable}, i.e. neither $x$ is preferrable over $y$, nor $y$ is preferrable over $x$): if $x$ is incomparable with $y$, and $y$ is incomparable with $z$, then $x$ is also incomparable with $z$.
  \end{itemize}

@@

  Construction of graph is based on a statistical arbitrage method.
  In short, an edge going from node $i$ to node $j$ with weight $w_{i,j}$ will be present in graph if and only if following happens: assets represented by nodes $i$ and $j$ have demonstrated similar behavior during the past period of time, but have suddenly diverged at the moment, as determined per statistical measures.
  Weight $w_{i,j}$ corresponds to the magnitude of this divergence.  
  A detailed description of procedure is given later in \ref{sub:creating-graph}.

%  The measure of this preference flow is determined by a statistical arbitrage method, and it corresponds to the magnitude of a pair of assets prices ratio going out of what is considered statistically confident range.
%

  The main idea of preference flow is that preference of a particular node should depend on amount of preference flowing in and out of it.
  This allows for some kind of generalized statistical arbitrage over multiple assets at a time.

@@

    \item Afterwards, we can calculate the reconstruction $\matr{f}^*$ of preference  flow by simply plugging back $\matr{\phi^*}$ into (\ref{eq:flowformula}):
    \begin{equation}
    \matr{f^*} = \matr{B} \matr{\phi^*}.
    \end{equation}
    The reconstructed preference flow $\matr{f^*}$ compared to the original preference flow $\matr{f}$ may even contain some new and/or lose some old edges.
    In addition, $\matr{B}$, $\matr{\phi^*}$, and $\matr{f^*}$ now describe a consistent graph $\graph{G}^*$.
    This allows now to define a consistency measure $\kappa$ as follows:
    \begin{equation}
    \label{eq:consistency}
    \kappa = \frac{\left \lVert \matr{f^*} \right \rVert}{\left \lVert \matr{f} \right \rVert}.
    \end{equation}
    Equation (\ref{eq:consistency}) represents the cosine of the angle between $\matr{f}$ and $\matr{f^*}$ in the column space of matrix $\matr{B}$.
    Consistency measure $\kappa$ tells us how consistent graph $\graph{G}$ was, compared to the $\graph{G}^*$.
    $\kappa$ ranges from 0 to 1, with 0 meaning full inconsistency (virtually unreachable), and 1 meaning full consistency.

@@

  \subsection{Creating the graph of preference flow}
  \label{sub:creating-graph}
  Using the obtained $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, and $d_{i,j}^{(t)}$ it is now possible to create a graph of preference flow among assets for each time step $t$.
  Considering one time step $t$, we find all such pairs of assets $(i,j)$ for which holds:  
  \begin{equation}
    \label{eq:thresh}
    \left| c_{i,j}^{(t)} - m_{i,j}^{(t)} \right| > \alpha \cdot d_{i,j}^{(t)},
  \end{equation}
  i.e. current log price difference is at least $\alpha$ deviations distant from mean value of the past time window.
  An example is shown on Fig. \ref{fig:devmag}.
%  Parameter $\alpha$ determines how many pairs of assets should constitute the graph at current time step.
  Afterwards, for each observed pair $(i,j)$ that exceeds the threshold we add into graph vertices $i$ and $j$, with a weighed edge of weight $w_{i,j}^{(t)}$ going from $i$ to $j$.
  Weight $w_{i,j}^{(t)}$ is obtained as:
  \begin{equation}
    \label{eq:weight}
    w_{i,j}^{(t)} = \left. \left(c_{i,j}^{(t)} - m_{i,j}^{(t)}\right) \middle/ d_{i,j}^{(t)} \right..
  \end{equation}

@@

  % \section*{APPENDIX}
